---
title: "Statistical Design Final Project"
author: "Russell Whitworth"
date: "2025-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#clear memory
rm(list = ls())
```

# Overview
Suppose we’d like to address the power for a new RCBD study with 6 treatments and anticipate the mean of the 4 treatments can be estimated as [80,83,86,89], and the block variance is estimated as 25, and the residual variance is estimated as 55.

## Load in libraries
```{r}
library(WebPower)
```

## Given Data
```{r}
treatment_means <- c(80, 83, 86, 89)   # Treatment means
block_variance <- 25                    # Block variance
residual_variance <- 55                 # Residual variance
sigma_squared <- block_variance + residual_variance # Total variance
```

## Calculating Effect Size (Cohen's ƒ)
```{r}
mean_treatment <- mean(treatment_means)
between_group_variance <- sum((treatment_means - mean_treatment)^2) / (length(treatment_means) - 1)
f_value <- sqrt(between_group_variance / sigma_squared) # Cohen's f (effect size)
f_value
```
This is considered a large effect size according to Cohen’s benchmarks (large ≥ 0.4).

## Power Analysis for Different Block Sizes
```{r}
# For 5 blocks
power_5 <- wp.anova(f = f_value, k = 6, n = 30, alpha = 0.05, power = NULL)$power

# For 10 blocks
power_10 <- wp.anova(f = f_value, k = 6, n = 60, alpha = 0.05, power = NULL)$power

# For 15 blocks
power_15 <- wp.anova(f = f_value, k = 6, n = 90, alpha = 0.05, power = NULL)$power

# For 20 blocks
power_20 <- wp.anova(f = f_value, k = 6, n = 120, alpha = 0.05, power = NULL)$power

# For 25 blocks
power_25 <- wp.anova(f = f_value, k = 6, n = 150, alpha = 0.05, power = NULL)$power

# For 30 blocks
power_30 <- wp.anova(f = f_value, k = 6, n = 180, alpha = 0.05, power = NULL)$power

# Display results as a data frame
power_results <- data.frame(
  Blocks = c(5, 10, 15, 20, 25, 30),
  Power = round(c(power_5, power_10, power_15, power_20, power_25, power_30), 4)
)

print(power_results)

```

```{r}
wp_target <- wp.anova(f = f_value, k = 6, power = 0.8, alpha = 0.05)
wp_target$n/6 # This is the number of blocks needed
```
13 blocks are needed for 0.8 power.